---
title: "persistencePits v7.0"
author: "Matthew E. Verbyla"
date: "6/16/2022"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

First, load the required packages.

```{r packages, message=FALSE}
library(ggplot2)
library(gridExtra)
library("PerformanceAnalytics")
library("Hmisc")
library("knitr")
library("bbmle")
```

## Load and explore data

Now, we will load and explore the persistence data, which are stored on the GWPP K2P Data Portal under the following dataset: http://data.waterpathogens.org/dataset/persistence-treatment. 

```{r dataLoad, echo=FALSE}
persist<-read.csv("http://data.waterpathogens.org/dataset/eda3c64c-479e-4177-869c-93b3dc247a10/resource/f99291ab-d536-4536-a146-083a07ea49b9/download/k2p_persistence.csv",header=T)
persist<-persist[persist$matrix=="Fecal sludge",]
persist$ln_removal<--persist$log10_reduction*log(10)
persist$additive <- as.character(persist$additive)
persist$additive <- replace(persist$additive,persist$additive=="None","a_None")
persist$additive <- as.factor(persist$additive)
persist$urine <- as.character(persist$urine)
```

The next chunk of code explores the data set to quantify the total number of data points, experiments, and the breakdown of lab-based studies, field-based studies, and the number of experiments done with different types of pathogen groups.

```{r dataExplore}
nrow(persist) # total number of data points
length(unique(persist$experiment_id)) # total number of experiments
length(persist[persist$location_comment=="Laboratory-based study",]$experiment_id) # number of data points from lab-based studies
length(unique(persist[persist$location_comment=="Laboratory-based study",]$experiment_id)) # number of lab-based experiments
length(persist[persist$location_comment!="Laboratory-based study",]$experiment_id) # number of data points from field-based studies
length(unique(persist[persist$location_comment!="Laboratory-based study",]$experiment_id)) # number of field-based experiments
length(unique(persist[persist$microbial_group=="Virus",]$experiment_id)) # number of experiments with viruses
length(unique(persist[persist$microbial_group=="Bacteria",]$experiment_id)) # number of experiments with bacteria
length(unique(persist[persist$microbial_group=="Protozoa",]$experiment_id)) # number of experiments with protozoa
length(unique(persist[persist$microbial_group=="Helminth",]$experiment_id)) # number of experiments with helminths
unique(persist[persist$microbial_group=="Virus",]$microorganism_common_name) # types of viruses studied
unique(persist[persist$microbial_group=="Bacteria",]$microorganism_common_name) # types of bacteria studied
unique(persist[persist$microbial_group=="Protozoa",]$microorganism_common_name) # types of protozoa studied
unique(persist[persist$microbial_group=="Helminth",]$microorganism_common_name) # types of helminths studied
```

## Calculate the decay rate coefficients

The next chunk of code sets up some variables that will be used in an upcoming "FOR" loop, which will be used to calculate the decay rate coefficients and other statistics associated with the fit of the log-linear and JM2 models.

```{r setUpForLoop}
N<-length(unique(persist$experiment_id)) # the total number of experiments
M<-length(persist$experiment_id) # the total number of data points
persist$ind<-NA
for(i in 1:M){
  # this creates a unique index variable to distinguish each independent experiment
  persist$ind[i]<-which(data.frame(unique(persist$experiment_id))==persist$experiment_id[i])
}
# here we create lots of NULL variables, which later we will populate values for in the upcoming "for" loop
k<-NULL;k1<-NULL;k2<-NULL;intrcpt<-NULL;group<-NULL;addit<-NULL;urine<-NULL;urea<-NULL;moist<-NULL;temp<-NULL;pH<-NULL;r2<-NULL;t99<-NULL;t99jm<-NULL;authors<-NULL;num<-NULL;aicLL<-NULL;aicJM<-NULL;experiment<-NULL;maxLRV<-NULL;maxTime<-NULL;mcn<-NULL;best<-NULL;sterr<-NULL;sigme<-NULL
par(mfrow=c(3,6)) #this creates the panel layout for the plots used for creating Figure S1
```

```{r forLoop, fig.show='hide', warning=FALSE}
for(i in 1:N){   #in this loop, we calculate the decay rate coefficients for the log-linear and JM2 models
  time<-persist[persist$ind==i,]$time_days   # get the time for the present experiment
  lnrv<-persist[persist$ind==i,]$ln_removal  # get the ln reduction for the present experiment
  log10_reduction<-persist[persist$ind==i,]$log10_reduction # get the log10 reduction for the present experiment
  # since we calculated the ln reduction, then equation gets algebraically rearranged like this:
  #         Ct = Co*exp(-k*t) 
  #         ln(Ct/Co) = -k*t
  
  # first, we fit the model like this to express the decay rate coefficient, k, on a LN scale
  fit<-lm(lnrv~time)  # lnrv = ln(Ct/Co), so our linear model is lnrv~time
  # then, we fit the model like this to create plots and calculate T99 values more easily later on
  fit2<-lm(log10_reduction~time)
  
  # next, we set up the JM2 model
  JM2<-function(k1,k2,sigma){
    R = lnrv - log(1/(1+exp(k1+k2*log(time)))) # this is the equation for the JM2 model
    R = suppressWarnings(dnorm(R,0,sigma,log=T)) # this is the standard deviation of the normally distributed difference between measured and modeled LRVs
    -sum(R)  # this is the negative log likelihood, which must be maximized (using the MLE method)
  }
  
  # in order to use MLE to find the best values for the JM2 model, we have to provide some halfway decent guesses for k1, k2, and sigma
  # through trial and error, we found that starting guesses of k1=-5, k2=1, and sigma=3 generally produced converging results for most experiments; the following are exceptions
  if(fit$coefficients[2]<=-10){k1guess=-fit$coefficients[2];k2guess=-fit$coefficients[2];siguess=3}else{
    if(fit$coefficients[2]>=-0.0001){k1guess=-5;k2guess=0.01;siguess=2}else{
      if(i==48|i==181|i==186){k1guess=-0.1;k2guess=1;siguess=1}else{
        if(i==115){k1guess=4;k2guess=100;siguess=1}else{
          if(i==112|i==113|i==105|i==114){k1guess=5;k2guess=1;siguess=1}else{
            if(i==180){k1guess=1;k2guess=10;siguess=1}else{
              k1guess=-5;k2guess=1;siguess=3
            }
          }
        }
      }
    }
  }

  if(length(time)>3){ # it only calculates the k1, k2, and sigma values for the JM2 model if there are more than 3 time points in the experiment
    fit.jm<-mle2(minuslogl=JM2,start=list(k1=k1guess,k2=k2guess,sigma=siguess),optimizer="nlminb") # uses unbounded optimization using portable Fortran programs for numerical computation (PORT) routines 
    # Reference: Fox et al. (1978). “The PORT Mathematical Subroutine Library.” ACM Transactions on Mathematical Software, 4(2), 104–126. doi:10.1145/355780.355783
  }else{fit.jm<-NA}
  
  ## NEXT WE STORE THE VALUES OF THE VARIABLES AND STATISTICS WE WANT TO STORE FOR EACH EXPERIMENT
  # here, we store the basic information about the experiment (e.g., the design, operational, and environmental parameters)  
  authors[i]<-as.character(unique(persist[persist$ind==i,]$authors)) # the names of the authors of the paper where the experiment's data were published
  experiment[i]<-as.numeric(unique(persist[persist$ind==i,]$experiment_id)) # the unique experiment ID
  num[i]<-length(time) # number of time points in the experiment
  maxTime[i]<-max(time) # the maximum time point for which data were collected in the experiment
  maxLRV[i]<-as.numeric(max(persist[persist$ind==i,]$log10_reduction)) # the maximum log10 reduction value observed in the experiment
  group[i]<-as.character(unique(persist[persist$ind==i,]$microbial_group)) # the microbial group used in the experiment (viruses, bacteria, protozoa, helminths)
  mcn[i]<-as.character(persist[persist$ind==i,]$microorganism_common_name) # a more specific description of the microbial group
  addit[i]<-as.character(unique(persist[persist$ind==i,]$additive)) # the type of additive applied to the fecal sludge
  urine[i]<-as.character(unique(persist[persist$ind==i,]$urine)) # whether the experiment was done with feces only or feces mixed with urine (excreta)
  moist[i]<-as.numeric(median(persist[persist$ind==i,]$moisture_content_percent)) # the average percent moisture content of the fecal sludge during the experiment
  temp[i]<-as.numeric(median(persist[persist$ind==i,]$temperature_celsius)) # the average temperature of the fecal sludge during the experiment
  pH[i]<-as.numeric(median(persist[persist$ind==i,]$pH)) # the average pH of the fecal sludge during the experiment
  urea[i]<-as.character(unique(persist[persist$ind==i,]$urea)) # whether or not urea (or stored urine) was added to the fecal sludge during the experiment

  # here, we store the statistics for the fit to the log-linear model
  k[i]<-fit$coefficients[2] # the log-linear decay rate coefficient
  intrcpt[i]<-fit$coefficients[1] # the log-linear model Y-intercept
  out<-summary(fit); sterr[i]<-out$coefficients[1,2] # this is the standard error of the decay rate coefficient
  t99[i]<-(2-fit2$coefficients[1])/fit2$coefficients[2] # the predicted T99 value for the log-linear model
  r2[i]<-summary(fit)$r.squared # coefficient of determination for the log-linear model fit
  aicLL[i]<-AIC(fit) # the AIC value for the log-linear model fit
  
  # and here, we store the statistics for the fit to the JM2 model
  if(!is.na(fit.jm)){k1[i]<-coef(fit.jm)["k1"]} # the 1st JM2 model decay rate coefficient
  if(!is.na(fit.jm)){k2[i]<-coef(fit.jm)["k2"]} # the 2nd JM2 model decay rate coefficient
  if(!is.na(fit.jm)){sigme[i]<-coef(fit.jm)["sigma"]} # the sigma value for the fit of the JM2 model (measure of uncertainty)
  t99jm[i]<-exp((log(99)-k1[i])/k2[i]) # the predicted T99 value for the JM2 model
  if(length(time)>3){aicJM[i]<-AIC(fit.jm)}else{aicJM[i]<-NA} # the AIC value for the JM2 model fit
  
  # this stores text describing which of the two models had the better fit (Chick Log-Linear Model or JM2 Model), based on the AIC values calculated
  if(is.na(aicJM[i])){best[i]<-"Chick Model"}else{if(aicLL[i]<aicJM[i]){best[i]<-"Chick Model"}else{best[i]<-"JM2 Model"}}

  # this next chunk of script creates the plot for each panel of Figure S1 (new plot created for each experiment during each round of the FOR loop)
  par(mar=c(5,4,3.5,1)) # sets the margins so that the plot looks nice
  plot(time,-log10_reduction,ylim=c(-8,0),xlim=c(0,500),ylab="Log10 Reduction",xlab="Time (days)",main=paste(experiment[i],", ",group[i],", N=",num[i],", LRVmax=",round(maxLRV[i],2),"\n",addit[i],", ",urea[i],", ",round(r2[i],2),"\n",authors[i],sep=""),sub=best[i],cex.main=0.95)
  lines(seq(0,360),-fit2$coefficients[2]*seq(0,360)-fit2$coefficients[1],col="blue")
  if(length(time)>3){lines(seq(0,360),log10(1/(1+exp(coef(fit.jm)["k1"]+coef(fit.jm)["k2"]*log(seq(0,360))))),lty=2,col="red")}
}
# now we will create a data frame with all the values from all experiments
kPit<-data.frame(experiment=experiment,authors=authors,microbial_group=group,mcn=mcn,k=k,se.k=sterr,k1=k1,k2=k2,se.jm=sigme,int=intrcpt,aicLL=aicLL,aicJM=aicJM,num=num,additive=addit,urea=urea,urine=urine,moisture=moist,temp=temp,pH=pH,r2=r2,maxLRV=maxLRV,t99=t99,t99jm=t99jm,bestModel=best)
kPit$k<-round(kPit$k,4)
kPit$k1<-round(kPit$k1,2)
kPit$k2<-round(kPit$k2,2)
kPit.demo<-kPit[c(1,17,126,236),c("experiment","authors","mcn","k","k1","k2")]
kPit.demo
```

The previous chunk of code hid the plots that were created for Figure S1. Here is a demonstrative plot for experiment 144.

```{r plotFits, warning=FALSE, echo=FALSE}
# this script creates the plot for the panel of Figure S1 that corresponds with experiment 144 only
par(mar=c(5,4,3.5,1)) # sets the margins so that the plot looks nice
i=17
time<-persist[persist$ind==i,]$time_days   # get the time for the present experiment
lnrv<-persist[persist$ind==i,]$ln_removal  # get the ln reduction for the present experiment
log10_reduction<-persist[persist$ind==i,]$log10_reduction # get the log10 reduction for the present experiment
fit<-lm(lnrv~time)  # lnrv = ln(Ct/Co), so our linear model is lnrv~time
fit2<-lm(log10_reduction~time)
plot(time,-log10_reduction,ylim=c(-8,0),xlim=c(0,500),ylab="Log10 Reduction",xlab="Time (days)",main=paste(experiment[i],", ",group[i],", N=",num[i],", LRVmax=",round(maxLRV[i],2),"\n",addit[i],", ",urea[i],", ",round(r2[i],2),"\n",authors[i],sep=""),sub=best[i],cex.main=0.95)
lines(seq(0,360),-fit2$coefficients[2]*seq(0,360)-fit2$coefficients[1],col="blue")
if(length(time)>3){lines(seq(0,360),log10(1/(1+exp(k1[i]+k2[i]*log(seq(0,360))))),lty=2,col="red")}
```

Here is a summary of which model provided the better fit for each experiment

```{r}
kPit[,c("experiment","mcn","bestModel")]
```

## Produce figures and tables

This last chunk of code uses the data frame kPit (which has all of the results from fitting the models to data from each experiment) to produce figures and tables.

###Table 1

```{r Table 1, warning=FALSE}
# sample size for k-value calcs based on log linear, then sample size for t99 calcs based on log linear, then the k values, then the estimated T99 values based on the log linear model
data.frame(rbind(c(paste(length(kPit[kPit$microbial_group=="Virus",]$k)," (",length(kPit[kPit$microbial_group=="Virus"&kPit$maxLRV>=1,]$t99),")",sep=""), 
  -quantile(kPit[kPit$microbial_group=="Virus",]$k,c(0.5,0.95,0.05)),
  quantile(kPit[kPit$microbial_group=="Virus"&kPit$maxLRV>=1,]$t99,c(0.5,0.05,0.95))),
c(paste(length(kPit[kPit$microbial_group=="Bacteria",]$k)," (",length(kPit[kPit$microbial_group=="Bacteria"&kPit$maxLRV>=1,]$t99),")",sep=""), 
  -quantile(kPit[kPit$microbial_group=="Bacteria",]$k,c(0.5,0.95,0.05)),
  quantile(kPit[kPit$microbial_group=="Bacteria"&kPit$maxLRV>=1,]$t99,c(0.5,0.05,0.95))),
c(paste(length(kPit[kPit$microbial_group=="Protozoa",]$k)," (",length(kPit[kPit$microbial_group=="Protozoa"&kPit$maxLRV>=1,]$t99),")",sep=""), 
  -quantile(kPit[kPit$microbial_group=="Protozoa",]$k,c(0.5,0.95,0.05)),
  quantile(kPit[kPit$microbial_group=="Protozoa"&kPit$maxLRV>=1,]$t99,c(0.5,0.05,0.95))),
c(paste(length(kPit[kPit$microbial_group=="Helminth",]$k)," (",length(kPit[kPit$microbial_group=="Helminth"&kPit$maxLRV>=1,]$t99),")",sep=""), 
  -quantile(kPit[kPit$microbial_group=="Helminth",]$k,c(0.5,0.95,0.05)),
  quantile(kPit[kPit$microbial_group=="Helminth"&kPit$maxLRV>=1,]$t99,c(0.5,0.05,0.95)))
))
```

###Table 2

```{r Table 2, warning=FALSE}
data.frame(rbind(
  c(length(kPit[kPit$microbial_group=="Virus"&kPit$num>3&kPit$maxLRV>=1,]$t99jm),
    quantile(kPit[kPit$microbial_group=="Virus"&kPit$num>3&kPit$maxLRV>=1,]$k1,c(0.5,0.05,0.95),na.rm=T),
    quantile(kPit[kPit$microbial_group=="Virus"&kPit$num>3&kPit$maxLRV>=1,]$k2,c(0.5,0.05,0.95),na.rm=T),
    quantile(kPit[kPit$microbial_group=="Virus"&kPit$num>3&kPit$maxLRV>=1,]$t99jm,c(0.5,0.05,0.95),na.rm=T)),
  c(length(kPit[kPit$microbial_group=="Bacteria"&kPit$num>3&kPit$maxLRV>=1,]$t99jm),
    quantile(kPit[kPit$microbial_group=="Bacteria"&kPit$num>3&kPit$maxLRV>=1,]$k1,c(0.5,0.05,0.95),na.rm=T),
    quantile(kPit[kPit$microbial_group=="Bacteria"&kPit$num>3&kPit$maxLRV>=1,]$k2,c(0.5,0.05,0.95),na.rm=T),
    quantile(kPit[kPit$microbial_group=="Bacteria"&kPit$num>3&kPit$maxLRV>=1,]$t99jm,c(0.5,0.05,0.95),na.rm=T)),
  c(length(kPit[kPit$microbial_group=="Protozoa"&kPit$num>3&kPit$maxLRV>=1,]$t99jm),
    quantile(kPit[kPit$microbial_group=="Protozoa"&kPit$num>3&kPit$maxLRV>=1,]$k1,c(0.5,0.05,0.95),na.rm=T),
    quantile(kPit[kPit$microbial_group=="Protozoa"&kPit$num>3&kPit$maxLRV>=1,]$k2,c(0.5,0.05,0.95),na.rm=T),
    quantile(kPit[kPit$microbial_group=="Protozoa"&kPit$num>3&kPit$maxLRV>=1,]$t99jm,c(0.5,0.05,0.95),na.rm=T)),
  c(length(kPit[kPit$microbial_group=="Helminth"&kPit$num>3&kPit$maxLRV>=1,]$t99jm),
    quantile(kPit[kPit$microbial_group=="Helminth"&kPit$num>3&kPit$maxLRV>=1,]$k1,c(0.5,0.05,0.95),na.rm=T),
    quantile(kPit[kPit$microbial_group=="Helminth"&kPit$num>3&kPit$maxLRV>=1,]$k2,c(0.5,0.05,0.95),na.rm=T),
    quantile(kPit[kPit$microbial_group=="Helminth"&kPit$num>3&kPit$maxLRV>=1,]$t99jm,c(0.5,0.05,0.95),na.rm=T))
))
```

###Figure 2

```{r figure2, warning=FALSE}
persistExc<-persist[!persist$experiment_id %in% kPit$experiment,]
persistInc<-persist[persist$experiment_id %in% kPit$experiment,]

plotFun<-function(persistanceData){
  fitV <- lm(-log10_reduction ~ time_days, data = persistanceData[persistanceData$microbial_group=="Virus" & persistanceData$matrix=="Fecal sludge" & persistanceData$time_days<200,])
  new_df <- persistanceData[persistanceData$microbial_group=="Virus" & persistanceData$matrix=="Fecal sludge" & persistanceData$time_days<200,]   
  vi <- ggplot(new_df, aes(time_days, -log10_reduction)) +
    geom_point() +
    ylim(-8,0) + ylab("Log Reduction Value") + 
    xlim(0,150) + xlab("Time (days)") +
    ggtitle("Viruses") +
    theme(plot.title = element_text(hjust=0.5)) +
    theme_bw() #+ geom_smooth(method=lm, se=TRUE)
  fitB <- lm(-log10_reduction ~ time_days, data = persistanceData[persistanceData$microbial_group=="Bacteria" & persistanceData$matrix=="Fecal sludge" & persistanceData$time_days<200,])
  new_df <- persistanceData[persistanceData$microbial_group=="Bacteria" & persistanceData$matrix=="Fecal sludge" & persistanceData$time_days<200,]
  ba <- ggplot(new_df, aes(time_days, -log10_reduction))+
    geom_point() +
    ylim(-8,0) + ylab("Log Reduction Value") + 
    xlim(0,150) + xlab("Time (days)") +
    ggtitle("Bacteria") +
    theme(plot.title = element_text(hjust=0.5))+
    theme_bw()
  fitH <- lm(-log10_reduction ~ time_days, data = persistanceData[persistanceData$microbial_group=="Helminth" & persistanceData$matrix=="Fecal sludge" & persistanceData$time_days<200,])
  new_df <- persistanceData[persistanceData$microbial_group=="Helminth" & persistanceData$matrix=="Fecal sludge" & persistanceData$time_days<200,]
  he <- ggplot(new_df, aes(time_days, -log10_reduction))+
    geom_point() +
    ylim(-8,0) + ylab("Log Reduction Value") + 
    xlim(0,150) + xlab("Time (days)") +
    ggtitle("Helminths") +
    theme(plot.title = element_text(hjust=0.5))+
    theme_bw()
  fitP <- lm(-log10_reduction ~ time_days, data = persistanceData[persistanceData$microbial_group=="Protozoa" & persistanceData$matrix=="Fecal sludge" & persistanceData$time_days<200,])
  new_df <- persistanceData[persistanceData$microbial_group=="Protozoa" & persistanceData$matrix=="Fecal sludge" & persistanceData$time_days<200,]
  pr <- ggplot(new_df, aes(time_days, -log10_reduction))+
    geom_point() +
    ylim(-8,0) + ylab("Log Reduction Value") + 
    xlim(0,150) + xlab("Time (days)") +
    ggtitle("Protozoa") +
    theme(plot.title = element_text(hjust=0.5))+
    theme_bw()
grid.arrange(vi,ba,pr,he,nrow = 2,widths = c(2,2))
}
plotFun(persistInc) # Plots Figure 2
```

###Figure 3

```{r figure3}
par(mfrow=c(1,3))
myData<-kPit[kPit$temp<50&kPit$temp>0,c("tk","k1","tk2","urine","urea","additive")]
myData$urine<-factor(myData$urine)
myData[myData$additive=="None",]$additive <- "a_None"
a<-ggplot(myData,aes(urine,tk))+geom_boxplot(aes(fill=urine))+theme_bw()+scale_fill_grey(start=0,end=.9)
b<-ggplot(myData,aes(urea,tk))+geom_boxplot(aes(fill=urea))+theme_bw()+scale_fill_grey(start=0,end=.9)
c<-ggplot(myData,aes(additive,tk))+geom_boxplot(aes(fill=additive))+theme_bw()+scale_fill_grey(start=0,end=.9)
grid.arrange(a,b,c,nrow=1,widths=c(1,1,2))

a<-ggplot(myData,aes(urine,k1))+geom_boxplot(aes(fill=urine),outlier.shape=NA)+theme_bw()+scale_fill_grey(start=0,end=.9)+coord_cartesian(ylim=c(-100,50))
b<-ggplot(myData,aes(urea,k1))+geom_boxplot(aes(fill=urea),outlier.shape=NA)+theme_bw()+scale_fill_grey(start=0,end=.9)+coord_cartesian(ylim=c(-100,50))
c<-ggplot(myData,aes(additive,k1))+geom_boxplot(aes(fill=additive),outlier.shape=NA)+theme_bw()+scale_fill_grey(start=0,end=.9)+coord_cartesian(ylim=c(-100,50))
grid.arrange(a,b,c,nrow=1,widths=c(1,1,2))

a<-ggplot(myData,aes(urine,tk2))+geom_boxplot(aes(fill=urine),outlier.shape=NA)+theme_bw()+scale_fill_grey(start=0,end=.9)+coord_cartesian(ylim=c(1.5,2))
b<-ggplot(myData,aes(urea,tk2))+geom_boxplot(aes(fill=urea),outlier.shape=NA)+theme_bw()+scale_fill_grey(start=0,end=.9)+coord_cartesian(ylim=c(1.5,2))
c<-ggplot(myData,aes(additive,tk2))+geom_boxplot(aes(fill=additive),outlier.shape=NA)+theme_bw()+scale_fill_grey(start=0,end=.9)+coord_cartesian(ylim=c(1.5,2))
grid.arrange(a,b,c,nrow=1,widths=c(1,1,2))
```

###Table 1

```{r table1}
kPit$lk<-log(-kPit$k)
fit.final<-lm(lk~factor(microbial_group)+pH+temp+moisture+factor(urine)+factor(urea)+factor(additive),data=kPit[kPit$temp<50,]) #Regression
summary(fit.final)
```

###Figure S4

```{r figureS4, warning=FALSE}
kPit$tk<-log(-kPit$k)
kPit$tk1<-sqrt(log(kPit$k1+150))
kPit$tk2<-sqrt(log(kPit$k2+10))

par(mfrow=c(3,2),mar=c(4,4,2,2))
hist(kPit[kPit$temp<50&kPit$temp>0,]$k,breaks=100)
hist(kPit[kPit$temp<50&kPit$temp>0,]$tk,breaks=100)
hist(kPit[kPit$temp<50&kPit$temp>0,]$k1,breaks=100)
hist(kPit[kPit$temp<50&kPit$temp>0&kPit$k1>-20&kPit$k1<20,]$tk1,breaks=100)
hist(kPit[kPit$temp<50&kPit$temp>0,]$k2,breaks=100)
hist(kPit[kPit$temp<50&kPit$temp>0&kPit$k1<50,]$tk2,breaks=100)
```

###Figure S5

```{r figureS5}
myData<-kPit[kPit$temp<50&kPit$temp>0,c("microbial_group","k","k1","k2","urine","urea","additive")]
a<-ggplot(myData,aes(microbial_group,k))+geom_boxplot(aes(fill=microbial_group),outlier.shape=NA)+theme_bw()+scale_fill_grey(start=0,end=.9)+coord_cartesian(ylim=c(-15,1))
b<-ggplot(myData,aes(microbial_group,k1))+geom_boxplot(aes(fill=microbial_group),outlier.shape=NA)+theme_bw()+scale_fill_grey(start=0,end=.9)+coord_cartesian(ylim=c(-35,10))
c<-ggplot(myData,aes(microbial_group,k2))+geom_boxplot(aes(fill=microbial_group),outlier.shape=NA)+theme_bw()+scale_fill_grey(start=0,end=.9)+coord_cartesian(ylim=c(-10,30))
grid.arrange(a,b,c,nrow=1,widths=c(1,1,1))
```

###Figure S6

```{r figureS6, warning=FALSE}
myData<-kPit[kPit$temp<50&kPit$temp>0,c("tk","k1","tk2","temp","pH","moisture")]
names(myData)<-c("ln(k) (1/days)","k1","tk2","Temperature (°C)","pH","moisture")
par(mfrow=c(1,2))
chart.Correlation(kPit[,c("tk","k1","tk2","temp","pH","moisture")],histogram=TRUE)
chart.Correlation(myData,histogram=TRUE)
```

###Figure S7

```{r figureS7, warning=FALSE}
fit.final<-lm(lk~factor(microbial_group)+pH+temp+moisture+factor(urine)+factor(urea)+factor(additive),data=kPit)
plot(fit.final,which=1:6)
```
